Spatiotemporal Optimization of Cosmological Particel-Mesh Simulation

### 2023-01-12 docs/papers/sto/sto.tex
Draft: just a few sentences

\subsection{Snapshot Observable}

@@

\subsection{Spatial Optimization}

@@

Force sharpening
%
\begin{equation}
\frac{k_i}{k^2} \to \frac{k_i}{k^2}
  h(k_i; \vvartheta) g(k; \vvartheta) \prod_{j=1}^3 f(k_j; \vvartheta),
\end{equation}
%
where $f$, $g$, $h$ are neural networks, each approximating a nonlinear
function of the dimensionless features, e.g.,
%
\begin{equation}
g(k; \vvartheta) = g(k/k_P, kR_\tophat, \cdots; D, f, \cdots).
\end{equation}

\subsection{Temporal Optimization}

@@

The time-dependence of the discrete symplectic integrator mainly enters through
the drift and kick factors in \eqref{DFK}.
Besides, the input features of the SO NNs could also depend on time.
Therefore when we perform TO together with SO, the force evaluation in
\eqref{DFK} which contains SO also becomes time-dependent and needs to be
optimized.

\begin{table*}
\centering
\caption{Ranges of GADGET-4 and \pmwd\ configuration and cosmological
parameters.
Note that the grid ratio need next fast len to determine the mesh shape
for fast FFT.
Given the box size, the mesh shape determines the cell size.
The softening parameter gives the ratio of the comving softening length
to the mean particle spacing.
The curvature $\Omega_k$ is related to the separate universe simulation.
We sample parameters applicable to GADGET-4 during data generation, and
those applicable to \pmwd\ during training.
\textsuperscript\dag The box size is determined jointly by the \pmwd\
mesh shape and mesh cell size below, which are assumed to be sampled
independently.
In practice, we sample one conditioned on the other and the box size.
* Number of time steps from $a=1/16$ to $a=1$.
}
\label{tab:param}
\begin{tabular}{lcccr}
  \toprule
  applicability & parameter & distribution & range & sampling \\
  \midrule
  & $128^3$ white noise & $\mathcal{N}(\vzero, \vI)$ & 512 realizations & MC \\
  \cmidrule(lr){2-5}
  & 121 snapshots, $a$ & uniform & [1/16, 1] & deterministic \\
  \cmidrule(lr){2-5}
  & box size in Mpc\textsuperscript\dag & log-trapezoidal & [25.6, 2560) \\
  \cmidrule(lr){2-4}
  & snapshot offset, $\Delta\!a$ & uniform & [0, 1/128) \\
  \cmidrule(lr){2-4}
  GADGET-4 & $\As \times 10^9$ & log-uniform & [1, 4) \\
  \cmidrule(lr){2-4}
  \& \pmwd\ & $\ns$ & log-uniform & [0.75, 1.25) & RQMC \\
  \cmidrule(lr){2-4}
  & $\Omegam$ & log-uniform & [1/5, 1/2) & see \autoref{fig:sobol} \\
  \cmidrule(lr){2-4}
  & $\Omegab / \Omegam$ & log-uniform & [1/16, 1/2) \\
  \cmidrule(lr){2-4}
  & $\Omega_k / (1 - \Omega_k)$ & uniform & [-0.5, 0.5) \\
  \cmidrule(lr){2-4}
  & $h$ & log-uniform & [0.5, 1) \\
  \cmidrule(lr){1-4}
  GADGET-4 & softening ratio & log-uniform & [1/50, 1/20) \\
  \cmidrule(lr){1-5}
  & mesh shape & log-uniform & [128, 512] \\
  \cmidrule(lr){2-4}
  \pmwd\ & mesh cell size in Mpc & log-uniform & [0.2, 5] & MC \\
  \cmidrule(lr){2-4}
  & number of time steps* & log-uniform & [10, 1000] \\
  \bottomrule
\end{tabular}
\end{table*}

@@

>>>>

\subsection{Snapshot Observable}
\label{sec:snapobs}

@@

\subsection{Spatial Optimization}
\label{sec:so}

@@

In Fourier space, the gravitational force field reads
\begin{equation}
- i \vk \varphi(\vk) = i \frac32 \Omegam H_0^2 \,\frac{\vk}{k^2} \delta(\vk),
\end{equation}
Force sharpening
%
\begin{equation}
\frac{k_i}{k^2} \to \frac{k_i}{k^2}
  h(k_i; \vvartheta) g(k; \vvartheta) \prod_{j=1}^3 f(k_j; \vvartheta),
\end{equation}
%
where $f$, $g$, $h$ are neural networks, each approximating a nonlinear
function of the dimensionless features, e.g.,
%
\begin{equation}
g(k; \vvartheta) = g(k/k_P, kR_\tophat, \cdots; D, f, \cdots).
\end{equation}

Parameters in these neural networks are optimized during training.

\subsection{Temporal Optimization}
\label{sec:to}

@@

The time-dependence of the discrete symplectic integrator mainly enters through
the drift and kick factors in \eqref{DFK}.
Besides, the input features of the SO NNs introduced in \autoref{sec:so} could
also depend on time.
Therefore when we perform TO together with SO, the force evaluation in
\eqref{DFK} which contains SO also becomes time-dependent and needs to be
optimized.

\begin{table*}
\centering
\caption{Ranges of GADGET-4 and \pmwd\ configuration and cosmological
parameters.
Note that the grid ratio need next fast len to determine the mesh shape
for fast FFT.
Given the box size, the mesh shape determines the cell size.
The softening parameter gives the ratio of the comoving softening length
to the mean particle spacing.
The curvature $\Omega_k$ is related to the separate universe simulation.
We sample parameters applicable to GADGET-4 during data generation, and
those applicable to \pmwd\ during training.
\textsuperscript\dag The box size is determined jointly by the \pmwd\
mesh shape and mesh cell size below, which are assumed to be sampled
independently.
In practice, we sample one conditioned on the other and the box size.
* Number of time steps from $a=1/16$ to $a=1$.
}
\label{tab:param}
\begin{tabular}{lcccr}
  \toprule
  applicability & parameter & distribution & range & sampling \\
  \midrule
  & $128^3$ white noise & $\mathcal{N}(\vzero, \vI)$ & 512 realizations & MC \\
  \cmidrule(lr){2-5}
  & 121 snapshots, $a$ & uniform & [1/16, 1] & deterministic \\
  \cmidrule(lr){2-5}
  & box size in Mpc\textsuperscript\dag & log-trapezoidal & [25.6, 2560) \\
  \cmidrule(lr){2-4}
  & snapshot offset, $\Delta\!a$ & uniform & [0, 1/128) \\
  \cmidrule(lr){2-4}
  GADGET-4 & $\As \times 10^9$ & log-uniform & [1, 4) \\
  \cmidrule(lr){2-4}
  \& \pmwd\ & $\ns$ & log-uniform & [0.75, 1.25) & RQMC \\
  \cmidrule(lr){2-4}
  & $\Omegam$ & log-uniform & [1/5, 1/2) & see \autoref{fig:sobol} \\
  \cmidrule(lr){2-4}
  & $\Omegab / \Omegam$ & log-uniform & [1/16, 1/2) \\
  \cmidrule(lr){2-4}
  & $\Omega_k / (1 - \Omega_k)$ & uniform & [-0.5, 0.5) \\
  \cmidrule(lr){2-4}
  & $h$ & log-uniform & [0.5, 1) \\
  \cmidrule(lr){1-4}
  GADGET-4 & softening ratio & log-uniform & [1/50, 1/20) \\
  \cmidrule(lr){1-5}
  & mesh shape & log-uniform & [128, 512] \\
  \cmidrule(lr){2-4}
  \pmwd\ & mesh cell size in Mpc & log-uniform & [0.2, 5] & MC \\
  \cmidrule(lr){2-4}
  & number of time steps* & log-uniform & [10, 1000] \\
  \bottomrule
\end{tabular}
\end{table*}

@@

\subsection{Training}

Given 512 Gadget simulations and 121 output snapshots each, in total we have
61952 snapshots for training.
Following \autoref{tab:param}, given a snapshot from Gadget simulations, we
configure \pmwd\ with the same initial coniditions and cosmological parameters,
and randomly sample the mesh shape (or the mesh cell size) and the number of
time steps.
